Syntax and Semantics of Abstract Binding Trees

### 2015-11-13 paper.tex
[paper] adjust modes and edge styles
Clean up packages
Use tikz for inline arrows

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{agda} % this should come after ams*
\usepackage{catchfilebetweentags}
%\usepackage{fontspec}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{ifdraft}
\usepackage{ifthen}
%\usepackage{lualatex-math}
\usepackage{mathtools}
\usepackage{proof}
\usepackage{scalerel}
\usepackage{setspace}
\usepackage{stmaryrd}
\usepackage{supertabular}
\usepackage{tikz-cd}
%\usepackage{unicode-math}
\usepackage{url}
%\usepackage{xunicode}

@@

% fonts
\usepackage{pxfonts}
\usepackage{eulervm}
\usepackage{bbold}
\usepackage[
  activate={true,nocompatibility},
  kerning=true,
  spacing=true,
  tracking=true
]{microtype}
\microtypecontext{spacing=nonfrench}
% \setmainfont{Linux Libertine O}
% \setmathfont{Asana Math}

@@

\newcommand\Iso[2]{#1\mathrel{{{\cong}}}#2}
\newcommand\Cats{\mathbf{Cat}}
\newcommand\Discrete[1]{#1_{\equiv}}
\newcommand\Arr[1]{\mathbf{{#1}^{\rightarrow}}}
\newcommand\Sets{\mathbf{Set}}
\newcommand\SCtx{\mathbf{SCtx}}
\newcommand\Ctx{\mathbf{Ctx}}
\newcommand\Seq{\mathbf{Seq}}
\newcommand\HypCat{\mathbf{H}}

@@

Let $\Symbols$ be an infinite set of symbols. Let $\SymSets$ be the category of
finite subsets of $\Symbols$ and their injective maps; then the comma
construction $\Define{\SCtx}{\Comma{\SymSets}{\Discrete{\Sorts}}}$, with
$\Discrete{\Sorts}$ the discrete category on the set $\mathcal{S}$, is the
category of contexts of symbols, whose objects are finite sets of symbols $U$
and sort-assignments $\Of{\mathfrak{s}}{U\to\Sorts}$, and whose morphisms are
sort-preserving renamings; we will write $\Upsilon$ for a symbol context
$(U,\mathfrak{s})$.

Then, fix a covariant presheaf (copresheaf) of operators
$\Of{\Operators}{\SCtx\times\Discrete{\Arities}\to\Sets}$ such that the arrows
in $\SCtx$ lift to renamings of operators' parameters. Together, $\Sorts$ and
$\Operators$ form a \emph{signature} $\Define{\Sigma}{\Pair{\Sorts}{\Operators}}$.

By the Grothendieck construction
$\Iso{\Grothendieck{}{\Operators}}{\Comma{\mathbb{1}}{\Operators}}$ we have a category of
elements of $\Operators$ where objects are triples
$\Pair{\Pair{\Upsilon}{a}}{\vartheta}$, interpreted as
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$, and morphisms are paired
diagrams:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\Pair{\Upsilon}{a}}
      \arrow[r, "\Pair{\rho}{f}", swap]
    \& {\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
  \qquad\begin{tikzcd}[cramped, sep = small, ampersand replacement = \&]
    \& \mathbb{1}
      \arrow[dl, "\vartheta", swap]
      \arrow[dr, "\vartheta'"]
    \&
    \\
    {\Operators\Pair{\Upsilon}{a}}
      \arrow[rr, "\Operators\Pair{\rho}{f}", swap]
    \&
    \& {\Operators\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\]
We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and note that this judgment
enjoys the structural properties of weakening and exchange via functoriality of
$\Operators$. Operators $\vartheta_{0}, \ldots, \vartheta_{n}$ are defined by
specifying the fibers of
$\Of{\pi_{\Operators}}{\Grothendieck{}{\Operators}\to\SCtx\times\Discrete{\Arities}}$
in which they reside:
\begin{alignat*}{3}
  &\IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}&&\,\mathrel{\ {{:}}}\,&&\IMode{\SCtx\times\Discrete{\Arities}\to\Sets}\\
  &\IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}&&\,=\,\ && %
    \OMode{%
      \MkSet{
        \vartheta\in\Pair{\Upsilon}{a}
        \mid
        {\pi_{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}} =
        \Pair{\Upsilon}{a}
      }
    }
\end{alignat*}

@@

Declaring a new assignable consists in providing an initial value, and an
expression binding a symbol (which shall represent the assignable in scope).
Note that the functoriality of $\Operators$ guarantees for any renaming
$\Of{\varrho}{\Upsilon\to\Upsilon'}$ and $\Of{f}{a\to a'}$, a lifted map
$\Of{\FLift{\varrho}{f}}{\Operators\Pair{\Upsilon}{a}\to\Operators\Pair{\Upsilon'}{a'}}$
such that $\IsOperator{\Upsilon'}{\FLift{\varrho}{f}(\vartheta)}{a'}$ when
$\IsOperator{\Upsilon}{\vartheta}{a}$. Because the category
$\Discrete{\Arities}$ is discrete and has only identity arrows, through an abuse
of notation we will often write $\FLift{\varrho}{}$ to mean
$\FLift{\varrho}{\ArrId{}}$. In particular, the renaming
$\Upsilon,u\mapsto\Upsilon,v$ shall take $\mathtt{get}[u]$ to $\mathtt{get}[v]$.

@@

The only place that symbols appear in our calculus is as parameters to
operators (unlike variables, symbols are not terms). Therefore, the functorial
action of the operator copresheaf can be lifted into terms by recursion on their
structure, via a pair of judgments $\IsRenaming{\varrho}{M}{N}$ and
$\IsRenaming{\varrho}{E}{F}$, presupposing
$\Of{\varrho}{\Upsilon\to\Upsilon'}$, and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$
and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$ respectively:

@@

\begin{remark}
  A presheaf $P$ with a substitution monoidal structure $\Pair{\nu}{\varsigma}$
  also forms a \emph{relative monad}\footnote{Relative monads are skew-monoids
    in the skew-monoidal structure induced by a functor category
    $\mathbf{C}^\mathbf{J}$~\cite{altenkirch-chapman-uustalu:2015}. However,
    relative monads can also be understood without reference to monoids by
    viewing them as a generalization of \emph{extension systems} or
    \emph{Kleisli structures}. This latter presentation of monads, popular in
    computer science and logic, does not involve iteration of the monad
    functor.} on the functor $\Of{V}{\widehat{\HypCat}^\Sorts}$ of variables.
  Consequently, monoid multiplication $\varsigma$ can be replaced with extension
$\Of{\circledast}{\int_{\Tuple{\tau, (\Upsilon\parallel\Gamma), (\Upsilon'\parallel\Gamma')}}{{{P}_{\tau}(\Upsilon'\parallel\Gamma')}^{{V}_{\tau}(\Upsilon\parallel\Gamma)} \longrightarrow {{P}_{\tau}(\Upsilon'\parallel\Gamma')}^{{P}_{\tau}(\Upsilon\parallel\Gamma)}}}$.
\end{remark}

@@

For each signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we have an
endofunctor
$\Of{\SigFun{\Sigma}}{\widehat{\HypCat}^\Sorts\to\widehat{\HypCat}^\Sorts}$,
which is defined as follows:
\begin{align*}
  \IMode{{\SigFun{\Sigma}(X)}_{\tau}}
    &\triangleq
    \OMode{
      \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}
      \prod_{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau_i}\in\vec{v}}
      X_{\tau_i}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}
    }
\end{align*}

Then, a $\Sigma$-model is a $\Sigma$-monoid $P$ which is equipped with an initial
algebra $\Of{\alpha}{\SigFun{\Sigma}(P)\to P}$, which shall interpret
applications of each operator.

@@

Then, the interpretation of a term in a model $P$ is a map from its environment
to $P$:
\[%
  \Of{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
      M:\tau}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P\longrightarrow P_\tau}%
\]

Variables are interpreted by the map
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P$ which
projects them from the environment and embeds them into the model. Metavariables
are resolved by the map
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}
  : \tau}_P$ (where
$\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$) which
projects their interpretation from the environment and instantiates it via
substitution:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{P_\tau}}
    \&[8.0em] {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[rr, dashed, "\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P"]
      \ar{drr}[
        , rotate = -16.5
        , xshift = +1.5em
        , yshift = -0.2em
        , swap
      ]{{\nu}_{\Gamma} {\pi}_{3}}
      \ar{dl}[
        , rotate = +11.5
        , xshift = -2.75em
      ]{\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}}
      \arrow[l, dashed, swap, "\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}: \tau}_P"]
    \&[-2.0em]
    \&[+3.0em]{\IMode{P_\tau}}
    \\
    {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
      \arrow[u, "\IMode{\varsigma^\tau_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \&{}
    \&{}
    \& {\IMode{P^{[\Gamma]}}}
      \arrow[u, swap, "{\pi}_{x}"]
    \\
    {\IMode{P^{[\vec{\tau}]}}}
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[l, swap, "\Define{\psi\,}{\,\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}}"]
    \& {\IMode{S^{\{\Upsilon\}}}}
      \arrow[r, "\Define{\phi\,}{\,\Tuple{\pi_u}_{u\in\vec{u}}}"]
    \& {\IMode{S^{\{\vec{\sigma}\}}}}
  \end{tikzcd}
\]

Interpretation of operator applications is the most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed by
symbol collections; therefore, operators must pass through suitable renamings in
order to be used in the interpretation. Let us begin by constructing for each
operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the morphism
$\SemBrackets{\vartheta}_P$ which shall rename the parameters of the operator
using the environment:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    |[alias = Src]|
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, swap, "\IMode{\pi_2}"]
    \&
    {\IMode{S^{\{\Upsilon\}}}}
      \arrow[r, swap, "\IMode{\FLift{\pi_{(-)}}{}}"]
    \&[1.5em]
    |[alias = Tgt]|
    {\IMode{\Operators_a}}
    \arrow[from = Src, to = Tgt, bend left, dashed, "\IMode{\SemBrackets{\vartheta}_{P}}"]
  \end{tikzcd}
\]

@@

The construction of $\beta$ proceeds by renaming the parameters of the operator
$\vartheta$ and constructing the (bound) exponentiated arguments $\gamma$ of the
operator.
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
       \arrow[r, "\Define{\beta\,}{\,\Tuple{\SemBrackets{\vartheta}_P, \lambda\gamma}}"]
    \&[4.0em] {\IMode{%
       \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}%
       \prod_{\MkValence{\vec{\sigma}_i}{\vec{\tau}_i}{\tau_i}\in\vec{v}}%
       P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}%
    }}
  \end{tikzcd}
\]
Arguments $\lambda\IMode{{\gamma}_{i}}$ are the exponential transposes (curried
form) of the composites $\IMode{{\gamma}_{i}}$:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}}}
      \arrow[d, dashed, swap, "\IMode{\lambda{\gamma}_{i}}"]
    \&
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}
       \times\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}
    }
      \arrow[r, "\IMode{\Tuple{{\pi}_{[1,1]},{\phi}_{i},{\psi}_{i}}}"]
      \arrow[d, dashed, swap, "\IMode{{\gamma}_{i}}"]
    \&[3.5em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_{i}}_{P}}}
      \ar{dl}[
        , rotate = +19.25
        , xshift = -4.5em
        , yshift = -0.2em
      ]{\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_{i}\parallel\Gamma,\vec{\tau}_{i}\vdash M_i:\tau_i}_P}}
    \\[2.0em]
    {\IMode{{P}_{{\tau}_{i}}^{\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}}}
    \& {\IMode{P_{{\tau}_{i}}}}
    \&
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Pair{\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}}}{\,\Tuple{\pi_u\circ\pi_{[2,1]}}_{u\in\Dom{\vec{\sigma}_i}}}}"]
    \&[12.0em] {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}\\[-1.0em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Pair{\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}}}{\,\Tuple{\pi_x\circ\pi_{[2,1]}}_{x\in\Dom{\vec{\tau}_i}}}}"]
    \&[12.0em] {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}

>>>>

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{agda} % this should come after ams*
\usepackage{catchfilebetweentags}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{ifdraft}
\usepackage{ifthen}
\usepackage{mathtools}
\usepackage{proof}
\usepackage{scalerel}
\usepackage{setspace}
\usepackage{stmaryrd}
\usepackage{supertabular}
\usepackage{tikz-cd}
\usepackage{tikzpfeile}
\usepackage{url}

@@

% fonts
\usepackage{pxfonts}
\usepackage{eulervm}
\usepackage{bbold}
\usepackage[
  activate={true,nocompatibility},
  kerning=true,
  spacing=true,
  tracking=true
]{microtype}
\microtypecontext{spacing=nonfrench}

@@

\newcommand\Iso[2]{#1\mathrel{{{\cong}}}#2}
\newcommand\Cats{\mathbf{Cat}}
\newcommand\Discrete[1]{#1_{\equiv}}
\newcommand\Arr[1]{\mathbf{{#1}^{\ra}}}
\newcommand\Sets{\mathbf{Set}}
\newcommand\SCtx{\mathbf{SCtx}}
\newcommand\Ctx{\mathbf{Ctx}}
\newcommand\Seq{\mathbf{Seq}}
\newcommand\HypCat{\mathbf{H}}

@@

Let $\Symbols$ be an infinite set of symbols. Let $\SymSets$ be the category of
finite subsets of $\Symbols$ and their injective maps; then the comma
construction $\Define{\SCtx}{\Comma{\SymSets}{\Discrete{\Sorts}}}$, with
$\Discrete{\Sorts}$ the discrete category on the set $\mathcal{S}$, is the
category of contexts of symbols, whose objects are finite sets of symbols $U$
and sort-assignments $\Of{\mathfrak{s}}{U\ra\Sorts}$, and whose morphisms are
sort-preserving renamings; we will write $\Upsilon$ for a symbol context
$(U,\mathfrak{s})$.

Then, fix a covariant presheaf (copresheaf) of operators
$\Of{\Operators}{\SCtx\times\Discrete{\Arities}\ra\Sets}$ such that the arrows
in $\SCtx$ lift to renamings of operators' parameters. Together, $\Sorts$ and
$\Operators$ form a \emph{signature} $\Define{\Sigma}{\Pair{\Sorts}{\Operators}}$.

By the Grothendieck construction
$\Iso{\Grothendieck{}{\Operators}}{\Comma{\mathbb{1}}{\Operators}}$ we have a category of
elements of $\Operators$ where objects are triples
$\Pair{\Pair{\Upsilon}{a}}{\vartheta}$, interpreted as
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$, and morphisms are paired
diagrams:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\Pair{\Upsilon}{a}}
      \arrow[r, "\Pair{\rho}{f}", swap]
    \& {\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
  \qquad\begin{tikzcd}[cramped, sep = small, ampersand replacement = \&]
    \& \mathbb{1}
      \arrow[dl, "\vartheta", swap]
      \arrow[dr, "\vartheta'"]
    \&
    \\
    {\Operators\Pair{\Upsilon}{a}}
      \arrow[rr, "\Operators\Pair{\rho}{f}", swap]
    \&
    \& {\Operators\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\]
We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and note that this judgment
enjoys the structural properties of weakening and exchange via functoriality of
$\Operators$. Operators $\vartheta_{0}, \ldots, \vartheta_{n}$ are defined by
specifying the fibers of
$\Of{\pi_{\Operators}}{\Grothendieck{}{\Operators}\ra\SCtx\times\Discrete{\Arities}}$
in which they reside:
\begin{alignat*}{3}
  &\IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}&&\,\mathrel{\ {{:}}}\,&&\IMode{\SCtx\times\Discrete{\Arities}\ra\Sets}\\
  &\IMode{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}&&\,=\,\ && %
    \OMode{%
      \MkSet{
        \vartheta\in\Pair{\Upsilon}{a}
        \mid
        {\pi_{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}} =
        \Pair{\Upsilon}{a}
      }
    }
\end{alignat*}

@@

Declaring a new assignable consists in providing an initial value, and an
expression binding a symbol (which shall represent the assignable in scope).
Note that the functoriality of $\Operators$ guarantees for any renaming
$\Of{\varrho}{\Upsilon\ra\Upsilon'}$ and $\Of{f}{a\ra a'}$, a lifted map
$\Of{\FLift{\varrho}{f}}{\Operators\Pair{\Upsilon}{a}\ra\Operators\Pair{\Upsilon'}{a'}}$
such that $\IsOperator{\Upsilon'}{\FLift{\varrho}{f}(\vartheta)}{a'}$ when
$\IsOperator{\Upsilon}{\vartheta}{a}$. Because the category
$\Discrete{\Arities}$ is discrete and has only identity arrows, through an abuse
of notation we will often write $\FLift{\varrho}{}$ to mean
$\FLift{\varrho}{\ArrId{}}$. In particular, the renaming
$\Upsilon,u\mapsto\Upsilon,v$ shall take $\mathtt{get}[u]$ to $\mathtt{get}[v]$.

@@

The only place that symbols appear in our calculus is as parameters to
operators (unlike variables, symbols are not terms). Therefore, the functorial
action of the operator copresheaf can be lifted into terms by recursion on their
structure, via a pair of judgments $\IsRenaming{\varrho}{M}{N}$ and
$\IsRenaming{\varrho}{E}{F}$, presupposing
$\Of{\varrho}{\Upsilon\ra\Upsilon'}$, and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$
and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$ respectively:

@@

\begin{remark}
  A presheaf $P$ with a substitution monoidal structure $\Pair{\nu}{\varsigma}$
  also forms a \emph{relative monad}\footnote{Relative monads are skew-monoids
    in the skew-monoidal structure induced by a functor category
    $\mathbf{C}^\mathbf{J}$~\cite{altenkirch-chapman-uustalu:2015}. However,
    relative monads can also be understood without reference to monoids by
    viewing them as a generalization of \emph{extension systems} or
    \emph{Kleisli structures}. This latter presentation of monads, popular in
    computer science and logic, does not involve iteration of the monad
    functor.} on the functor $\Of{V}{\widehat{\HypCat}^\Sorts}$ of variables.
  Consequently, monoid multiplication $\varsigma$ can be replaced with extension
$\Of{\circledast}{\int_{\Tuple{\tau, (\Upsilon\parallel\Gamma), (\Upsilon'\parallel\Gamma')}}{{{P}_{\tau}(\Upsilon'\parallel\Gamma')}^{{V}_{\tau}(\Upsilon\parallel\Gamma)} \ra {{P}_{\tau}(\Upsilon'\parallel\Gamma')}^{{P}_{\tau}(\Upsilon\parallel\Gamma)}}}$.
\end{remark}

@@

For each signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we have an
endofunctor
$\Of{\SigFun{\Sigma}}{\widehat{\HypCat}^\Sorts\ra\widehat{\HypCat}^\Sorts}$,
which is defined as follows:
\begin{align*}
  \IMode{{\SigFun{\Sigma}(X)}_{\tau}}
    &\triangleq
    \OMode{
      \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}
      \prod_{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau_i}\in\vec{v}}
      X_{\tau_i}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}
    }
\end{align*}

Then, a $\Sigma$-model is a $\Sigma$-monoid $P$ which is equipped with an initial
algebra $\Of{\alpha}{\SigFun{\Sigma}(P)\ra P}$, which shall interpret
applications of each operator.

@@

Then, the interpretation of a term in a model $P$ is a map from its environment
to $P$:
\[%
  \Of{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
      M:\tau}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P\ra P_\tau}%
\]

Variables are interpreted by the map
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P$ which
projects them from the environment and embeds them into the model. Metavariables
are resolved by the map
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}
  : \tau}_P$ (where
$\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$) which
projects their interpretation from the environment and instantiates it via
substitution:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{P_\tau}}
    \&[8.0em] {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[rr, dashed, "\OMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P}"]
      \ar{drr}[
        , rotate = -16.5
        , xshift = +1.5em
        , yshift = -0.2em
        , swap
      ]{{\nu}_{\Gamma} {\pi}_{3}}
      \ar{dl}[
        , rotate = +11.5
        , xshift = -2.75em
      ]{\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}}
      \arrow[l, dashed, swap, "\OMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}: \tau}_P}"]
    \&[-2.0em]
    \&[+3.0em]{\IMode{P_\tau}}
    \\
    {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
      \arrow[u, "\IMode{\varsigma^\tau_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \&{}
    \&{}
    \& {\IMode{P^{[\Gamma]}}}
      \arrow[u, swap, "{\pi}_{x}"]
    \\
    {\IMode{P^{[\vec{\tau}]}}}
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[l, dashed, swap, "\Define{\psi\,}{\,\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}}"]
    \& {\IMode{S^{\{\Upsilon\}}}}
      \arrow[r, dashed, "\Define{\phi\,}{\,\Tuple{\pi_u}_{u\in\vec{u}}}"]
    \& {\IMode{S^{\{\vec{\sigma}\}}}}
  \end{tikzcd}
\]

Interpretation of operator applications is the most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed by
symbol collections; therefore, operators must pass through suitable renamings in
order to be used in the interpretation. Let us begin by constructing for each
operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the morphism
$\SemBrackets{\vartheta}_P$ which shall rename the parameters of the operator
using the environment:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    |[alias = Src]|
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, swap, "\IMode{\pi_2}"]
    \&
    {\IMode{S^{\{\Upsilon\}}}}
      \arrow[r, swap, "\IMode{\FLift{\pi_{(-)}}{}}"]
    \&[1.5em]
    |[alias = Tgt]|
    {\IMode{\Operators_a}}
    \arrow[from = Src, to = Tgt, bend left, dashed, "\OMode{\SemBrackets{\vartheta}_{P}}"]
  \end{tikzcd}
\]

@@

The construction of $\beta$ proceeds by renaming the parameters of the operator
$\vartheta$ and constructing the (bound) exponentiated arguments $\gamma$ of the
operator.
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
       \arrow[r, dashed, "\Define{\beta\,}{\,\Tuple{\SemBrackets{\vartheta}_P, \lambda\gamma}}"]
    \&[4.0em] {\IMode{%
       \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}%
       \prod_{\MkValence{\vec{\sigma}_i}{\vec{\tau}_i}{\tau_i}\in\vec{v}}%
       P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}%
    }}
  \end{tikzcd}
\]
Arguments $\lambda\IMode{{\gamma}_{i}}$ are the exponential transposes (curried
form) of the composites $\IMode{{\gamma}_{i}}$:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}}}
      \arrow[d, dashed, swap, "\OMode{\lambda{\gamma}_{i}}"]
    \&
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}
       \times\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}
    }
      \arrow[r, "\IMode{\Tuple{{\pi}_{[1,1]},{\phi}_{i},{\psi}_{i}}}"]
      \arrow[d, dashed, swap, "\OMode{{\gamma}_{i}}"]
    \&[3.5em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_{i}}_{P}}}
      \ar{dl}[
        , rotate = +19.25
        , xshift = -4.5em
        , yshift = -0.2em
      ]{\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_{i}\parallel\Gamma,\vec{\tau}_{i}\vdash M_i:\tau_i}_P}}
    \\[2.0em]
    {\IMode{{P}_{{\tau}_{i}}^{\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}}}
    \& {\IMode{P_{{\tau}_{i}}}}
    \&
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, dashed, "\Define{\phi_i\,}{\,\Pair{\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}}}{\,\Tuple{\pi_u\circ\pi_{[2,1]}}_{u\in\Dom{\vec{\sigma}_i}}}}"]
    \&[12.0em] {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}\\[-1.0em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, dashed, "\Define{\psi_i\,}{\,\Pair{\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}}}{\,\Tuple{\pi_x\circ\pi_{[2,1]}}_{x\in\Dom{\vec{\tau}_i}}}}"]
    \&[12.0em] {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}
